---
format:
  gfm:
    default-image-extension: ""
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# econioread

<!-- badges: start -->
<!-- badges: end -->

econioread provides a step-by-step interface for reading input-output
tables from `csv`, `xls`, and `xlsx` files into
[econio](https://github.com/UchidaMizuki/econio) input-output table
objects. It handles the messy parts of importing published input-output
tables: locating and importing raw spreadsheet cells, building headers,
and identifying regions, sector names, and sector types, for both
regional and multiregional tables.

## Installation

You can install the development version of econioread from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("UchidaMizuki/econioread")
```

## Example

`io_table_reader()` builds a step-by-step reader for a single
input-output table file. Each `io_table_read_*()` step is applied in
turn with the pipe, and `end_step()` finishes the pipeline and returns
the `econio` input-output table.

The example below reads a typical published table whose first row is a
title, whose first two header rows and columns hold sector codes and
sector names, and whose sector blocks end with "total" rows and columns
such as `Total intermediate inputs` and `Total output`:

```{r}
#| label: example
#| eval: false
library(econioread)

io_table_reader("input-output-table.xlsx") |>
  io_table_read_cells(
    sheets = 1,
    rows_exclude = 1
  ) |>
  io_table_read_headers(
    input_names = c("input_sector_code", "input_sector_name"),
    output_names = c("output_sector_code", "output_sector_name")
  ) |>
  io_table_read_sector_names(
    input_sector_name_glue = "{input_sector_code}_{input_sector_name}",
    output_sector_name_glue = "{output_sector_code}_{output_sector_name}"
  ) |>
  io_table_read_sector_types(
    import_type = "competitive_import",
    industry_total_pattern = "Total intermediate",
    value_added_total_pattern = "Total value added",
    final_demand_total_pattern = "Total final demand",
    export_pattern = "Exports",
    export_total_pattern = "Total exports",
    import_pattern = "\\(less\\) Imports",
    import_total_pattern = "\\(less\\) Total imports",
    total_pattern = "Total output"
  ) |>
  io_table_read_data(value_scale = 1e6) |>
  end_step()
```

The steps work as follows:

- `io_table_read_cells()` imports the raw cells of the file.
  `rows_range`/`cols_range` and `rows_exclude`/`cols_exclude` trim
  rows and columns, such as titles and notes, that are not part of the
  table.
- `io_table_read_headers()` peels off header rows and columns into
  named columns. Input headers are taken from the `"left"` direction
  and output headers from `"up"` by default; other
  [unpivotr](https://nacnudus.github.io/unpivotr/) directions can be
  given as names, e.g. `c(up = "input_sector_code")`.
- `io_table_read_sector_names()` combines header columns into a single
  sector name with [glue](https://glue.tidyverse.org) patterns.
- `io_table_read_sector_types()` classifies sectors into industry,
  value-added, final-demand, export, import, and total sectors.
  `*_total_pattern` arguments locate the "total" row or column of each
  block and classify the sectors before it, while `*_pattern` regular
  expressions match sector names directly. `import_type` (either
  `"competitive_import"` or `"noncompetitive_import"`) controls which
  sector types appear on the input and output axes.
- `io_table_read_data()` parses the cell values, applies
  `value_scale`, and builds the input-output table with
  `econio::io_table_regional()` or `econio::io_table_multiregional()`.
  `total_tolerance` and `check_axes = FALSE` accommodate published
  tables whose subtotals contain rounding drift.
- `end_step()` returns the finished table.

## Custom steps

Published tables often need ad hoc cleaning between steps. Wrap any
data-frame function with `as_step()` (re-exported from the
[adverbial](https://github.com/UchidaMizuki/adverbial) package) to
insert it into the pipeline:

```{r}
#| label: custom-steps
#| eval: false
io_table_reader("iotable.xls") |>
  io_table_read_cells(sheets = 1) |>
  io_table_read_headers(
    input_names = c("input_sector_code", "input_sector_name"),
    output_names = c("output_sector_code", "output_sector_name")
  ) |>
  as_step(dplyr::mutate)(
    dplyr::across(
      c(input_sector_code, output_sector_code),
      \(x) stringr::str_remove(x, "^[:upper:]")
    )
  ) |>
  io_table_read_sector_names(
    input_sector_name_glue = "{input_sector_code}_{input_sector_name}",
    output_sector_name_glue = "{output_sector_code}_{output_sector_name}"
  ) |>
  as_step(dplyr::filter)(
    !stringr::str_detect(input_sector_name, "Subtotal$"),
    !stringr::str_detect(output_sector_name, "Subtotal$")
  ) |>
  io_table_read_sector_types(...) |>
  io_table_read_data(value_scale = 1e6) |>
  end_step()
```

## Multiregional tables

For multiregional tables, pass `region_type = "multiregional"` to
`io_table_reader()`, which adds an `io_table_read_regions()` step to
extract input and output regions before the sector names and types are
identified. In a multiregional table, the headers carry region codes
and names in addition to sector codes and names:

```{r}
#| label: example-multiregional
#| eval: false
io_table_reader(
  "multiregional-input-output-table.xlsx",
  region_type = "multiregional"
) |>
  io_table_read_cells(
    sheets = 1,
    rows_exclude = 1:4
  ) |>
  io_table_read_headers(
    input_names = c(
      "input_region_code",
      "input_region_name",
      "input_sector_code",
      "input_sector_name"
    ),
    output_names = c(
      "output_region_code",
      "output_region_name",
      "output_sector_code",
      "output_sector_name"
    )
  ) |>
  io_table_read_regions(
    input_region_glue = "{input_region_code}_{input_region_name}",
    output_region_glue = "{output_region_code}_{output_region_name}"
  ) |>
  io_table_read_sector_names(
    input_sector_name_glue = "{input_sector_code}_{input_sector_name}",
    output_sector_name_glue = "{output_sector_code}_{output_sector_name}"
  ) |>
  io_table_read_sector_types(
    import_type = "competitive_import",
    industry_total_pattern = "Total intermediate",
    value_added_total_pattern = "Total value added",
    final_demand_total_pattern = "Total final demand",
    export_pattern = "Exports",
    import_pattern = "\\(less\\) Imports",
    total_pattern = "Total output"
  ) |>
  io_table_read_data(value_scale = 1e6) |>
  end_step()
```
